---
title: "HMSC 1st model runs - KERGUELEN "
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---

The purpose here is to run a first series of models for Kerguelen and explore the results I extracted using the HMSC pipeline. Here I show the results for a model using 4 chains, 250 samples, and a thinning of 1000.


## Model inputs

```{r,echo = FALSE}

load("ker/models/S2_fit_models_thin_1000_samples_250_chains_4.Rdata")

```

```{r}
# Y = community pres/abs matrix of 1956 sites * 19 species (all but Limosella australis that was too rare)
# XData = the environmental matrix :
XData <- models$`presence-absence model`$XData
head(XData)

```

accum_prec = Accumulated precipitation amount over 1 year (bio12 CHELSA), 0.008333 resolution (~1 km)

mean_temp = mean temperature downscaled to ~30m, based on mean annual daily mean air temperatures averaged over 1 year (bio1 CHELSA), from 1981-2010

pente et expostion = terrain variables that come from field measures in the TAAF

jour mois annee = field sampling dates for pres/abs data


```{r}
# phylogenetic tree for Crozet. I kept the phylo.maker one.
phylotree <- ape::read.tree("../data/traits_trees/phylomaker_tree")
phylotree$tip.label <- gsub("_", " ", phylotree$tip.label)

# need to remove Notogrammitis crassior for Kerguelen
phylo_ker <- ape::drop.tip(phylotree, tip = "Notogrammitis crassior")
plot(phylo_ker)

```

Still no traits, and there wil be none from TRY ==> not enough traits measured across all plant species.


HMSC also needs the followong : 

```{r}

# STUDY DESIGN
studyDesign = data.frame(site=XData$numero_observation, id=XData$id)

# RANDOM EFFECT STRUCTURE, HERE Site (hierarchical study design)
rL.site = Hmsc::HmscRandomLevel(units = levels(studyDesign$site))
str(rL.site)

```

Not too sure what this does in detail, but it has default parameters to estimate the random effects. I didn't alter this.

```{r}
# and optionally id, if we are interested in species associations at that level
rL.id = Hmsc::HmscRandomLevel(units = levels(studyDesign$id))
str(rL.site)
```
Same structure as the site random level, except this one is supposed to be much finer, at the observation scale.



## Model structure 

```{r, eval=FALSE}
# REGRESSION MODEL FOR ENVIRONMENTAL COVARIATES.
XFormula = ~ mean_temp + accum_prec + pente + exposition

# REGRESSION MODEL FOR TRAITS : none.

# PRESENCE-ABSENCE MODEL FOR INDIVIDUAL SPECIES (COMMON ONLY)
m = Hmsc::Hmsc(Y=Y, XData = XData,  XFormula = XFormula,
         # TrData = TrData, TrFormula = TrFormula,
         phyloTree = phylo_ker,
         distr="probit",
         studyDesign = studyDesign, ranLevels=list(site=rL.site, id=rL.id))
```

## Model fit

input parameters : 

* nChains = 4

* nParallel = nChains

* samples=250, thin=1000


```{r, eval=FALSE}
m = Hmsc::sampleMcmc(m, samples = samples, thin=thin,
               adaptNf=rep(ceiling(0.4*samples*thin),m$nr), 
               transient = ceiling(0.5*samples*thin),
               nChains = nChains,
               nParallel = nParallel) 
```


## Results

See pdf called "ker/results/parameter_estimates_ex2.pdf").

* Variance partitioning plot: The corresponding values are in "ker/results/parameter_estimates_VP_presence_absence_model.csv"

There's also this : "ker/results/parameter_estimates_VP_R2T_Beta.csv" : it looks at the beta estimates for each level of the environmental factor variables, but it didn't work (NAs). Not sure why but doesn't matter too much for us I think.

* Beta plot : the posterior beta estimates for each plant * covariate. Corresponding values in "results/parameter_estimates_Beta_ presence-absence model.xls"

* Omega plots : associations between species at the site and id level.


## Evaluate model fit

I'm not sure I did this right. 

nfolds = NULL #Default: two-fold cross-validation. There's a place in the code where I think we're supposed to specify the column (variable) over which the fold should be done. I didn't..

It's all stored in the "ker/models/MF_thin_1000_samples_250_chains_4_nfolds_2.RData", and the results are in "ker/results/model_fit_nfolds_2.pdf". But I don't understand these plots.


## Make predictitons

See the "ker/results/predictions.pdf". 

























